Visualizing Project Evolution Through Abstract Syntax Tree Analysis

### 2016-07-11 paper/document.tex
Almost there...

Several visualizations tools have been created to understand how a software system changes over time. Wu \textit{et al.}~\cite{wu2004a} explored punctuated changes of software repositories with a tool that generates an Evolution Spectrograph. In this visualization, each file has a row in the chart and each cell represents a section of time where an incoming or outgoing dependency can be changed. The colour is determined by the frequency of change. This work was continued to show file directories and display the cardinality of a developer's commits in Wu \textit{et al.}~\cite{wu2004}. The cardinality of a commit is the number of files or subsystems it affects in a system, indicating the scope of the change. Other tools also look at changes in code dependencies~\cite{6980224} or files~\cite{6650525, 7332421, voinea2006} over time. Alcocer \textit{et al.}~\cite{6650523} create a Performance Evolution Blueprint, which displays the changes in performance over time as a project evolves. Ogawa and Ma created code\_swarm~\cite{codeswarm}, a generative art visualization which presents the progressive evolution of file changes made by authors in a source control repository over time. Gource~\cite{gource} is a similar visualization, showing an animated force-directed tree of the file hierarchy as authors make changes to files in the source code repository over time. Ens \textit{et al.} created ChronoTwigger to visualize the co-evolution of source and test files~\cite{6980223}. This allowed testing practices to be more closely examined. Anslow \textit{et al.}~\cite{sourcevis} created SourceVis, a suite of software visualizations including visualizations for semantically-aware project evolution. SourceVis displays the evolution of Java source code metrics such as number of packages, classes, methods, and fields.

@@

There is also research on how developers use software repositories to collaborate. Wagstrom \textit{et al.}~\cite{Patrick:Wagstrom:2012} explored the roles that developers take in networked, social development environments like GitHub. These roles were defined by their level of contribution as well as the types of issues they dealt with. They found that developers will take on multiple roles in a project and will sometimes fulfill the same role across different projects. Lee \textit{et al.}~\cite{lee2013} developed a tool for visualizing the branching structure of Git repositories. This was used to analyze concurrent workflows in popular open source repositories. Elsen also created a visualization for branches in Git, with the ability to expand branches to view the directory structure of a project at that point~\cite{6650522}.   Shrestha \textit{et al.} made a visualization for the geographic location of a commit and~\cite{6650532} and Minelli \textit{et al.}~\cite{6980226} looked at the real time changes developers made to a repository.

@@

\begin{description}
\item [declaration:]
  Stating that use of a field or variable of a given type. For example declaration of a \texttt{String} type variable might look as follows:
  \\
  \begin{adjustwidth}{2.5em}{0pt}
  \texttt{String myString;} \\
  \end{adjustwidth}
\item [invocation:]
  A call site for a method that belongs to a type. For example an invocation of the trim method of a String type might look as follows:
  \\
  \begin{adjustwidth}{2.5em}{0pt}
  \texttt{myString.trim();} \\
  \end{adjustwidth}
\end{description}

@@

The major downside of AST diffs over a method like line count is that ASTs are more expensive to compute. Our experiments were run on a Intel i7 with 16GB of RAM and an SSD. Calculating the AST diffs varied from an hour or two for smaller projects to a couple of days for very large projects. For example ANTLR4 took about 8 hours while Apache BookKeeper only took a couple of hours. The good news for ongoing projects is that AST diffs only need to be computed for new commits; TypeV can resume calculating diffs and augment statistics at any time.

@@

One issue with analyzing source code repositories is that authors can commit from a variety of different computers which can have different emails and usernames attached to their Git configuration. This may result in one author appearing as several different authors in the \texttt{git log}. There is no perfect way to deal with this problem; however, we added an interface for managing authors (Figure~\ref{fig:author-merging}. A user can manually specify an author's various aliases using a simple dropdown list. All statistics attributed to an author's various aliases are then remapped as belonging to one canonical author as specified by the user. The author managing view also provides a facility for selecting and deselecting authors shown in the timeline. Thus, visualization users can focus on the changes made by a particular author or a group of authors, rather than having to view the changes made by all authors.

@@

To determine if type coverage yields different information than file coverage, we analyzed four open source Java projects: ANTLR4, Apache BookKeeper, Apache Curator and Apache Tika (Table~\ref{tab:project-summary}). We chose ANTLR4 because it is a large Java project with a rich history of commits. ANTLR4 is also a member of the Qualitas Corpus~\cite{qualitas}, a curated collection of Java systems curated for empirical software engineering research. We then collected three smaller projects from the Apache foundation, as they host a large number of well-curated open source Java projects. These particular Apache projects were chosen for their clear development practices and available documentation. We counted the file coverage and type coverage on a commit-by-commit basis for each author. That is, each datum represents a commit with three dimensions: file coverage, type coverage, and its commit date. Data is sorted in ascending order of commit date. Table~\ref{tab:summary} lists a summary of the statistics measured.

We displayed the density distributions for two different projects: ANTLR4 (Figure~\ref{fig:antlr4-density}) and Apache Tika (Figure~\ref{fig:tika-density}). These plots show the distribution of the amount of commits that have X\% of type and file coverage. For example, Figure~\ref{fig:tika-density} shows that there are many commits with 50\% type coverage and while there are relatively few commits with 50\% file coverage. This shows that even though the type and file coverage plots may have a similar shape, the commits are showing different levels of information.

@@

% Discuss user study

@@

The timeline visualization presented in this paper (Section~\ref{sec:viz}) is intended to be a practical tool that enables a software developer's investigation of a project's evolution. Our timeline view can be seen as an evolution of simple visualizations such as GitHub's code frequency~\cite{github-graphs}. GitHub's graph is static, however it allows one to see an entire project's history at a glance. TypeV improves this by enabling one to see a particular \emph{type's} history at a glance. Given the greater density of information available, we made our timeline interactive such that a user is able to drill-down and explore the data at their own pace. Using interaction, TypeV not only shows what types have changed in project, but makes it possible to see \emph{why} a type has changed. We believe conventionality is a strong point in a developer's goal to make interesting discoveries in a project's evolution, but this has to be confirmed in user studies.

Bug triaging for large open source projects is a very time consuming task~\cite{badashian2015}, with hundreds  of bugs being reported daily. Each bug must be processed and assigned to a developer that has the experience required to fix the bug. In order to be effective, the triager has to have knowledge of what each developer has worked on as well as the parts of the code that may be affected by the bugs. With TypeV, the triager is able to directly see what types a developer has worked with over time. This is important information for triaging since the type causing the bug, if known, will have a list of associated developers who have recently used that type and can be assigned the bug. The commit messages for the affected types are also visible, providing greater context about how the type has been used. The assigned developer can also use type information to locate other pieces of code affected by a bug and know which developers have been using the affected types.

@@

>>>>

Several visualization tools have been created to understand how a software system changes over time. Wu \textit{et al.}~\cite{wu2004a} explored punctuated changes of software repositories with a tool that generates an Evolution Spectrograph. In this visualization, each file has a row in the chart and each cell represents a section of time where an incoming or outgoing dependency can be changed. The colour is determined by the frequency of change. This work was continued to show file directories and display the cardinality of a developer's commits in subsequent paper by Wu \textit{et al.}~\cite{wu2004}. The cardinality of a commit is the number of files or subsystems it affects in a system, indicating the scope of the change. Other tools also look at changes in code dependencies~\cite{6980224} or files~\cite{6650525, 7332421, voinea2006} over time. Alcocer \textit{et al.}~\cite{6650523} created a Performance Evolution Blueprint, which displays the changes in performance over time as a project evolves. Ogawa and Ma created code\_swarm~\cite{codeswarm}, a generative art visualization which presents the progressive evolution of file changes made by authors in a source control repository over time. Gource~\cite{gource} is a similar visualization, showing an animated force-directed tree of the file hierarchy as authors make changes to files in the source code repository over time. Ens \textit{et al.} created ChronoTwigger to visualize the co-evolution of source and test files~\cite{6980223}. This allowed testing practices to be more closely examined. Anslow \textit{et al.}~\cite{sourcevis} created SourceVis, a suite of software visualizations including visualizations for semantically-aware project evolution. SourceVis displays the evolution of Java source code metrics such as number of packages, classes, methods, and fields.

@@

There is also research on how developers use software repositories to collaborate. Wagstrom \textit{et al.}~\cite{Patrick:Wagstrom:2012} explored the roles that developers take in networked, social development environments like GitHub. These roles were defined by their level of contribution as well as the types of issues they dealt with. They found that developers will take on multiple roles in a project and will sometimes fulfill the same role across different projects. Lee \textit{et al.}~\cite{lee2013} developed a tool for visualizing the branching structure of Git repositories. This was used to analyze concurrent workflows in popular open source repositories. Elsen also created a visualization for branches in Git, with the ability to expand branches to view the directory structure of a project at that point~\cite{6650522}.   Shrestha \textit{et al.} made a visualization for the geographic location of a commit~\cite{6650532} and Minelli \textit{et al.}~\cite{6980226} looked at the real time changes developers made to a repository.

@@

\begin{description}
\item [declaration:]
  Stating the use of a field or variable of a given type. For example declaration of a \texttt{String} type variable might look as follows:
  \\
  \begin{adjustwidth}{2.5em}{0pt}
  \texttt{String myString;} \\
  \end{adjustwidth}
\item [invocation:]
  A call site for a method that belongs to a type. For example an invocation of the trim method of a String type might look as follows:
  \\
  \begin{adjustwidth}{2.5em}{0pt}
  \texttt{myString.trim();} \\
  \end{adjustwidth}
\end{description}

@@

The major downside of AST diffs over a method like line count is that ASTs are more expensive to compute. Our experiments were run on an Intel i7 with 16GB of RAM and an SSD. Calculating the AST diffs varied from an hour or two for smaller projects to a couple of days for very large projects. For example ANTLR4 took about 8 hours while Apache BookKeeper only took a couple of hours. The good news for ongoing projects is that AST diffs only need to be computed for new commits; TypeV can resume calculating diffs and augment statistics at any time.

@@

One issue with analyzing source code repositories is that authors can commit from a variety of different computers which can have different emails and usernames attached to their Git configuration. This may result in one author appearing as several different authors in the \texttt{git log}. There is no perfect way to deal with this problem; however, we added an interface for managing authors (Figure~\ref{fig:author-merging}). A user can manually specify an author's various aliases using a simple dropdown list. All statistics attributed to an author's various aliases are then remapped as belonging to one canonical author as specified by the user. The author managing view also provides a facility for selecting and deselecting authors shown in the timeline. Thus, visualization users can focus on the changes made by a particular author or a group of authors, rather than having to view the changes made by all authors.

@@

To determine if type coverage yields different information than file coverage, we analyzed four open source Java projects: ANTLR4, Apache BookKeeper, Apache Curator and Apache Tika (Table~\ref{tab:project-summary}). We chose ANTLR4 because it is a large Java project with a rich history of commits. ANTLR4 is also a member of the Qualitas Corpus~\cite{qualitas}, a curated collection of Java systems curated for empirical software engineering research. We then collected three smaller projects from the Apache foundation, as they host a large number of well-curated open source Java projects. These particular Apache projects were chosen for their clear development practices and easily available documentation. We counted the file coverage and type coverage on a commit-by-commit basis for each author. That is, each datum represents a commit with three dimensions: file coverage, type coverage, and its commit date. Data is sorted in ascending order of commit date. Table~\ref{tab:summary} lists a summary of the statistics measured.

We displayed the density distributions for two different projects: ANTLR4 (Figure~\ref{fig:antlr4-density}) and Apache Tika (Figure~\ref{fig:tika-density}). These plots were produced using R and show the distribution of the amount of commits that have X\% of type and file coverage. The red plot represents file coverage and the blue plot represents type coverage. The x-axis shows the percent of the coverage and the y-axis shows the number of commits which have the similar percent of coverage. For example, Figure~\ref{fig:tika-density} shows that there are many commits with 50\% type coverage and relatively few commits with 50\% file coverage. Also notice how in Figure~\ref{fig:antlr4-density} the type coverage is pushed to the right, meaning commits usually touch a higher percentage of types than files. So if we only looked at files touched we might be missing important changes in the project. Figures~\ref{fig:antlr4-density} and ~\ref{fig:tika-density} show that even though the type and file coverage plots may have a similar shape, the commits are showing different levels of information.

@@

@@

The timeline visualization presented in this paper (Section~\ref{sec:viz}) is intended to be a practical tool that enables users to investigate the evolution of a project. Our timeline view can be seen as an evolution of simple visualizations such as GitHub's code frequency~\cite{github-graphs}. GitHub's graph is static, however it allows one to see an entire project's history at a glance. TypeV improves this by enabling one to see a particular \emph{type's} history at a glance. Given the greater density of information available, we made our timeline interactive such that a user is able to drill-down and explore the data at their own pace. Using interaction, TypeV not only shows what types have changed in a project, but makes it possible to see \emph{why} a type has changed. We believe conventionality is a strong point in a developer's goal to make interesting discoveries in a project's evolution, but this has to be confirmed in user studies.

TypeV can also be useful for bug triaging. Bug triaging for large open source projects is a very time consuming task~\cite{badashian2015}, with hundreds  of bugs being reported daily. Each bug must be processed and assigned to a developer that has the experience required to fix the bug. In order to be effective, the triager has to have knowledge of what each developer has worked on as well as the parts of the code that may be affected by the bugs. With TypeV, the triager is able to directly see what types a developer has worked with over time. This is important information for triaging since the type causing the bug, if known, will have a list of associated developers who have recently used that type and can be assigned the bug. The commit messages for the affected types are also visible, providing greater context about how the type has been used. The assigned developer can also use type information to locate other pieces of code affected by a bug and know which developers have been using the affected types.

@@

A user study is needed to test the usefulness of TypeV. Some questions that need to be answered are: Can users new to a project easily find when major changes took place and, more importantly, what was being changed? Given a specific time period, can users find what was being worked on? The study should test TypeV against a line-based visualization such as GitHub’s graphs~\cite{github-graphs} and a file-based visualization such as code\_swarm~\cite{codeswarm}.
